Examining Programming Students’ Interaction with Generative AI: A Metacognitive Perspective

### 2024-10-03 CSEDU2025/Paper.tex
update paper

\keywords{Generative Artificial Intelligence, Metacognition, Programming Education, Chatbots}

\abstract{This study investigates the interaction patterns of programming
students with a generative AI-powered chatbot (CharlieBot) in the context of a
Java programming course. Specifically, it explores how students utilize the bot,
enhanced with Retrieval-Augmented Generation (RAG), along with two metacognitive
strategies: interleavBaha24Sun24ing and spacing. The methodology analyzed and
classified student conversations, focusing on types of interactions such as
conceptual questions and debugging help. The results indicate that most
interactions are focused on active learning strategies, although the deliberate use
of interleaving and spacing is limited. The study highlights the need for more
guidance on the strategic use of AI tools to improve learning outcomes.}

@@

Recent advances in Generative Artificial Intelligence (GenAI) have opened up
new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receive feedback, and obtain
detailed explanations.

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes towards the use of GenAI in
teaching and learning, noting that it improves the depth of their thinking and
understanding. Furthermore, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement. Besides that, it was also
observed that ChatGPT enables students to regulate their learning pace
\citep{Baha24} and can potentially support students’ self-regulated learning,
especially for those with previous technical and disciplinary knowledge
\citep{Xia23}.

@@

The primary objective of this study is to analyze the interactions between
students in a Java programming course and an educational bot powered by
ChatGPT 3.5, enhanced with Retrieval-Augmented Generation (RAG) capabilities,
within the context of two Metacognitive strategies: Interleaving and Spacing.
To achieve this objective, we formulate three research questions:

\begin{enumerate}
  \item What was the distribution pattern of the classified messages?
  \item Was there a space between the student prompts? If so, which category promotes the largest spacing between submissions?
  \item Do students' conversations with the bot alternate between practical and theoretical approaches, thereby promoting interleaving?
\end{enumerate}

@@

However, it is crucial to consider how GenAI tools impact students in
various demographic groups, academic disciplines, cultural backgrounds, and
levels of previous experience \citep{catalan21} \citep{neo22}. Consequently, this
study is limited to a specific group of students and focuses on the use of a
single GenAI tool. As a result, the findings may not be generalizable to other
populations or tools.

@@

% The advancements in GenAI and its use in programming courses have raised
% possibilities and concerns among educators and researchers. The study of
% \cite{Yilmaz23} found that the main benefit of using ChatGPT in programming
% learning is to provide providing fast and mostly correct answers to questions,
% improving thinking skills, facilitating debugging, and increasing
% self-confidence. In contrast, the study the same appoint limitations such ask
% laziness, being unable to answer some questions, or giving incomplete/incorrect
% answers, causing professional anxiety in students.

% From a cognitive perspective, \cite{Lo24} shown that students were able to
% learn from ChatGPT, which increased their understanding and achievement.
% However, concerns were raised that the growing use of AI tools might lead to a
% decline in critical thinking among students. Teachers and stakeholders should
% continue to investigate pedagogical approaches that leverage ChatGPT to enhance
% students' understanding and critical thinking. For example, teachers can
% instruct students to fact-check and validate information produced by ChatGPT.

% Examined students from different demographic groups using AI tools like
% ChatGPT, comparing perceptions and impacts among these groups. The results
% showed that AI tools had a positive impact on academic performance, especially
% for students with less prior experience.

% Critical thinking is a fundamental skill for students to develop, as it enables
% them to analyze, evaluate, and synthesize information. This skill is essential
% for problem-solving and decision-making, making it a crucial aspect of
% educational success, specifically in programming courses. However, the use of AI
% tools like ChatGPT has raised concerns about the potential impact on students'
% critical thinking abilities \cite{Murillo23} \cite{cai23} \cite{chan23}.

% In the other hand, the study of \cite{zhang24} suggested that incorporating
% ChatGPT could stimulate students' critical thinking, probably because they
% had to assess the correctness of ChatGPT's answers.

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.

@@

The use of generative artificial intelligence tools can support the important
mechanism of learning regulation by students \citep{larsen18}, since these
technologies frequently generate automatic and often accurate responses
\cite{Puryear22}, it becomes essential for students to use these tools more
strategically to promote conscious and active learning.

The Theory of Metacognition addresses awareness of mental processes.
\cite{flavell79} proposed a model of metacognitive monitoring that includes four
interrelated phenomena: Metacognitive Knowledge, Metacognitive Experience,
 Metacognitive Goals, and Metacognitive Actions. These processes do not occur
in isolation, but they influence each other, altering cognitive progress over
time.

@@

@@

CharlieBot is an educational bot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) capabilities to improve the precision of
responses provided to students \cite{Sun24}. The RAG feature was developed using
materials from a MOOC, including text, exercises, video transcripts, and other
resources.

@@

Since most of the messages are located in categories where students are
notoriously engaged in being more active in their learning (conceptual question
and problem solving help), corroborating the findings of \cite{Ghimire24} which
showed that the questions are also localized in the same categories.

>>>>

\keywords{
  Programming Education, Generative Artificial Intelligence, Chatbots and
  Metacognition
}

\abstract{
This study examines the interactions of programming students with a chatbot,
CharlieBot, powered by OpenAI's GPT-3.5 and enhanced with the
Retrieval-Augmented Generation (RAG) technique, within the context of a Java
programming course. It investigates how students use the bot while employing two
metacognitive strategies: interleaving and spacing. Using the logs of student
interactions with the bot, the methodology involved classifying and analyzing
student prompts into eight distinct categories. The findings reveal that while
most interactions emphasize active learning strategies, students demonstrate
limited conscious application of interleaving and spacing. This study
underscores the necessity for increased guidance on the strategic use of AI
tools to enhance learning outcomes.
}

@@

Recent advancements in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \citep{Peng23}
\citep{Pandey24} and education \citep{Puryear22}. In programming courses,
students can employ GenAI tools to enhance their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists by
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \citep{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and attempt to foster a deeper comprehension of programming concepts
\citep{Labadze23}.

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes towards the use of GenAI in
teaching and learning, noting that it may improve the depth of their thinking
and understanding. Furthermore, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement \citep{Callejo24}. Besides that,
it was also observed that ChatGPT enables students to regulate their learning
pace \citep{Baha24} and can potentially support students’ self-regulated
learning, especially for those with previous technical and disciplinary
knowledge \citep{Xia23}.

@@

In response to the growing need for deeper insights into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational bot powered by gpt-3.5 from OpenAI, enhanced with
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of the metacognitive strategies
employed by the students. To achieve this objective, we formulate three research
questions:

\begin{itemize}
  \item RQ1 - What was the distribution pattern of the classified messages?
  \item RQ2 - Was there a space between the student prompts? If so, which
  category promotes the largest spacing between submissions?
  \item RQ3 - Do students' conversations with the bot alternate between
  practical and theoretical approaches, thereby promoting interleaving?
\end{itemize}

@@

It is important to consider how GenAI tools impact students in
various demographic groups, academic disciplines, cultural backgrounds, and
levels of previous experience \citep{catalan21} \citep{neo22}. Consequently,
this study is limited to a specific group of students and focuses on the use of
a single GenAI tool. As a result, the findings may not be generalizable to other
populations or tools.

This paper is organized as follows: Section 2 presents the theoretical
framework, Section 3 describes the material and methods used in the study,
Section 4 presents the results and discussion, and Section 5 concludes the
paper and outlines future work.

@@

This section presents the theoretical framework that underpins the study. First,
we introduce the concept of metacognition used in this study. Next, we discuss
related work that explores the use of metacognitive strategies and scaffolds
in educational settings.

@@

Given that generative AI often produces automatic and highly accurate responses
\citep{Puryear22}, it is essential for students to use these tools within an
active learning process to enhance their learning outcomes. This active
learning process can be further understood through the lens of Metacognition, which
focuses on the awareness of one’s mental processes. \cite{flavell79} proposed a
model of metacognitive monitoring that includes four interrelated phenomena:
Metacognitive Knowledge, Metacognitive Experience, Metacognitive Goals, and
Metacognitive Actions. These processes do not occur in isolation, but they
influence each other, altering cognitive progress over time.

@@

Spacing and interleaving are two metacognitive strategies that have been shown to
enhance learning outcomes. Spacing refers to the practice of distributing study
sessions over time, which has been shown to improve long-term retention and
understanding of the material \citep{Carvalho20}. Interleaving involves mixing
different topics or problems within a single learning session, which has been
shown to enhance long-term learning and the application of knowledge in other
situations \citep{Rivers21}.

Studies on the use of metacognitive strategies indicate that these practices can
be applied to teaching the conscious use of generative AI. \cite{Zheng19}
investigated the effects of metacognitive scaffolds on group behavior,
performance, and cognitive load in computer-supported collaborative
environments. The results showed positive impacts on metacognitive
behavior and group performance without increasing cognitive load.

\cite{LiWei23} proposed a metacognition-based collaborative approach to enhance
performance in collaborative programming. The findings indicated that this
approach influenced computational thinking, critical thinking, and
metacognitive awareness.

Similarly, \cite{Wang23} explored how metacognitive instruction affects
computational thinking, critical thinking, and metacognitive skills in
collaborative programming, observing improvements in problem-solving and
collaboration.

These studies demonstrate the relevance of metacognition in enhancing
performance and skills in collaborative environments without imposing additional
cognitive load. Based on this, it is possible to suggest that metacognitive
strategies may also be useful in teaching a more conscious and critical use of
generative AI tools.

@@

CharlieBot is an educational bot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \cite{Sun24}. The RAG feature was developed using
materials from a MOOC, including text, exercises, video transcripts, and other
resources.

@@

Since most of the messages are located in categories where students are
notoriously engaged in being more active in their learning (Conceptual Question
and Debugging Help), corroborating the findings of \cite{Ghimire24} which
showed that the questions are also localized in the same categories.
